---
title: 'Exploring Territoriality as a Multivariate Phenotype'
author: "Gladiana Spitz"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, warning=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
par(mfrow=c(4,4),
    mar = c(2, 2, 2, 2))
library(tidyverse)
library(krsp)
library(lubridate)
library(sf)
library(spatstat)
library(cartogram)
library(adehabitatHR)
library(ggfortify)
library(ggplot2)
library(viridis)
library(lme4)
library(MCMCglmm)
library(igraph)
library(gridExtra)
library(pedtricks)
library(pedigree)

select = dplyr::select

knitr::opts_chunk$set(echo = TRUE)
con <- krsp_connect(host = "krsp.cepb5cjvqban.us-east-2.rds.amazonaws.com",
                    dbname ="krsp",
                    username = Sys.getenv("krsp_user"),
                    password = Sys.getenv("krsp_password")
)
con_suppl <- krsp_connect(host = "krsp.cepb5cjvqban.us-east-2.rds.amazonaws.com",
                          dbname ="krsp_suppl",
                          username = Sys.getenv("krsp_user"),
                          password = Sys.getenv("krsp_password")
)
# Script written by me to take KRSP locx and y coordinates and convert them to UTMs
# Newest update only uses sf to avoid retiring packages
source("locs_to_UTMs.R")
# script to pull tree cone counts
source("~/McAdam Lab/cone_count_summaries.R")
# script to pull densities
source("~/McAdam Lab/density.R")
# script to form pedigree
source("~/McAdam Lab/create_pedigree.R")
```

# How do we represent territoriality in red squirrels?

Territoriality is traditionally defined as the exclusive use of space to maintain access to a resource. Extensive data collection and early research in squirrels shows that individuals in environments like our population typically defend some amount of space around a larder cache of resources (midden).

## Gathering Phenotype Components

|Component|units|data tables|psuedocode or script file|
|---------|-----|-----------|----------|
|Territory Size|$ha$|census, trees, grid_stakes, behaviour|TerritorySize.Rmd|
| | | | |
|Probability of vocalization|ratio|trapping, census|RattleProbability.Rmd|
| | | | |
|Intrusion_Rate|ratio|trapping, census|IntrusionRate.Rmd|
| | | | |
|Number of total cones|count|flastall2, historic_squirrel_ids|CacheSize.Rmd|

# Prep Pedigree

```{r}
pedstat <- ped_stats(krsp_pedigree)
plot(pedstat)
summary(pedstat)


ped <- as.data.frame(krsp_pedigree)

ord <- orderPed(ped)
ped <- ped[order(ord),]

names(ped) <- c("animal","sire","dam")

Ainv <- inverseA(pedigree)$Ainv
```


# Loading in Territorial Phenotypes

```{r}
load("cache_size.rdata")

cache_sizes <- cache_sizes %>%
  mutate(animal = squirrel_id) %>%
  filter(!is.na(Grid),
         !is.na(Sex),
         !is.na(std_density),
         !is.na(std_local.density),
         !is.na(std_age))

load("IntrusionRate.rdata")

intruders_model <- intruders_model %>%
  mutate(animal = squirrel_id)

load("RattleProbability.rdata")

rattle_probability_final <- rattle_probability_final %>%
  filter(!is.na(Grid),
         !is.na(Sex),
         !is.na(std_density),
         !is.na(std_julian),
         !is.na(std_age)) %>%
  mutate(animal = squirrel_id)

load("TerritorySize.rdata")

data <- data %>%
  mutate(squirrel_id = as.factor(squirrel_id),
         animal = squirrel_id) %>%
  as.data.frame()
```

# What underlies variation in territoriality?

I aim to answer this in four parts. First I will use univariate phenotypic models that account for variation between individuals in each of the four territorial traits. Then I will use univariate animal models to account for differences in genetic and phenotypic components of the variation in each of the traits. I will repeat the building process from phenotypic to animal using a multivariate model.

The covariates will be grid, sex, year, spring grid density, local density in the fall. 

The random effects will be squirrel_id and year for phenotypic models. I will add animal and then dam id for animal models. 

Territory size and cache size are log transformed and treated as normally distributed. Rattle probability and intrusion rate are binary and should have a binomial error distribution. I will use parameter expanded priors as suggested by Jarrod Hadfield.

## Univariate Phenotypic Models

Set the prior

```{r}
PhenoPrior <- list(G=list(G1=list(V = diag(1), nu = 1, 
                                  alpha.mu = 0, alpha.V = diag(1)*1000),
                          G2=list(V = diag(1), nu = 1, 
                                  alpha.mu = 0, alpha.V = diag(1)*1000)),
                   R=list(V = diag(1) , nu = 1))

PhenoPrior_bi <- list(G=list(G1=list(V = diag(1), nu = 1, 
                                  alpha.mu = 0, alpha.V = diag(1)*1000),
                          G2=list(V = diag(1), nu = 1, 
                                  alpha.mu = 0, alpha.V = diag(1)*1000)),
                   R=list(V = diag(1) , fix = 1))
```

### Territory Size

```{r, fig.height=8, fig.width=8}
TSModel1 <- MCMCglmm(area_ha_log ~ Grid + Sex + std_density,
                     random = ~squirrel_id + year.f,
                     prior = PhenoPrior,
                     family = "gaussian",
                     data = data,
                     nitt = 250000, burnin = 50000, thin = 200,  verbose=F)

plot(TSModel1)
summary(TSModel1)

posterior.mode(TSModel1$VCV)
TSRepeatability <- TSModel1$VCV[,1] / 
  (TSModel1$VCV[,1] + TSModel1$VCV[,2] + 
     TSModel1$VCV[,3])
hist(TSRepeatability)
posterior.mode(TSRepeatability)
```

### Rattle Count

```{r, fig.height=8, fig.width=8}
RattleModel1 <- MCMCglmm(rattle_binary ~ Grid + Sex + spl(std_density, k=5) + 
                           spl(std_julian, k=3) + std_age,
                         random = ~squirrel_id + Year.f,
                         prior = PhenoPrior_bi,
                         family = "threshold",
                         data = rattle_probability_final, 
                         nitt = 250000, burnin = 50000, thin = 200, verbose=T)

summary(RattleModel1)
plot(RattleModel1)

posterior.mode(RattleModel1$VCV)
RattleRepeatability <- RattleModel1$VCV[,1] / 
  (RattleModel1$VCV[,1] + RattleModel1$VCV[,2] + RattleModel1$VCV[,3])
hist(RattleRepeatability)
posterior.mode(RattleRepeatability)
```

### Intrusion Count

```{r, fig.height=8, fig.width=8}
IntrusionModel1 <- MCMCglmm(intruded ~ Grid + Sex + spl(std_density, k=5) + 
                           spl(std_julian, k=5) + std_age + std_local.density,
                            random = ~squirrel_id + Year.f,
                            prior = PhenoPrior_bi,
                            family = "threshold",
                            data = intruders_model, 
                            nitt = 250000, burnin = 50000, thin = 200, verbose=F)
plot(IntrusionModel1)
posterior.mode(IntrusionModel1$VCV)
IntrusionRepeatability <- IntrusionModel1$VCV[,1] / 
  (IntrusionModel1$VCV[,1] + IntrusionModel1$VCV[,2] + IntrusionModel1$VCV[,3])
hist(IntrusionRepeatability)
posterior.mode(IntrusionRepeatability)
summary(IntrusionModel1)
```

### Cache Size

```{r, fig.height=8, fig.width=8}
CacheSizeModel1 <- MCMCglmm(cache_size_lastyear ~ Grid + Sex + std_density + 
                              std_age + I(std_age^2) + std_local.density,
                            random = ~squirrel_id + Year.f,
                            prior = PhenoPrior,
                            family = "gaussian",
                            data = cache_sizes, 
                            nitt = 250000, burnin = 50000, thin = 200, verbose=F)
plot(CacheSizeModel1)
posterior.mode(CacheSizeModel1$VCV)
CacheSizeRepeatability <- CacheSizeModel1$VCV[,1] / 
  (CacheSizeModel1$VCV[,1] + CacheSizeModel1$VCV[,2] + CacheSizeModel1$VCV[,3])
hist(CacheSizeRepeatability)
posterior.mode(CacheSizeRepeatability)
summary(CacheSizeModel1)
```

## Univariate Animal Models

Set the prior

```{r}
GenoPrior <- list(G=list(G1=list(V = diag(1), nu = 1, 
                                 alpha.mu = 0, alpha.V = diag(1)*1000),
                         G2=list(V = diag(1), nu = 1, 
                                 alpha.mu = 0, alpha.V = diag(1)*1000),
                         G3=list(V = diag(1), nu = 1, 
                                 alpha.mu = 0, alpha.V = diag(1)*1000)),
                  R=list(V = diag(1) , nu = 1))

GenoPrior_bi <- list(G=list(G1=list(V = diag(1), nu = 1, 
                                 alpha.mu = 0, alpha.V = diag(1)*1000),
                         G2=list(V = diag(1), nu = 1, 
                                 alpha.mu = 0, alpha.V = diag(1)*1000),
                         G3=list(V = diag(1), nu = 1, 
                                 alpha.mu = 0, alpha.V = diag(1)*1000)),
                  R=list(V = diag(1) , fix = 1))
```

### TS Area

```{r, fig.height=8, fig.width=8}

TSModel2 <- MCMCglmm(area_ha_log ~ Grid + Sex + std_density,
                     random = ~squirrel_id + year.f + animal,
                     prior = GenoPrior,
                     pedigree = Ainv,
                     family = "gaussian",
                     data = data,
                     nitt = 2500000, burnin = 500000, thin = 2000,  verbose=F)

summary(TSModel2)
plot(TSModel2)

TSRepeatability <- TSModel2$VCV[,1] / 
  (TSModel2$VCV[,1] + TSModel2$VCV[,2] + 
     TSModel2$VCV[,3] + TSModel2$VCV[,4])
hist(TSRepeatability)
posterior.mode(TSRepeatability)

TSHerit <- TSModel2$VCV[,3] / 
  (TSModel2$VCV[,1] + TSModel2$VCV[,2] + 
     TSModel2$VCV[,3] + TSModel2$VCV[,4])
hist(TSHerit)
posterior.mode(TSHerit)
```

### Rattle Count

```{r, fig.height=8, fig.width=8}
RattleModel2 <- MCMCglmm(rattle_binary ~ Grid + Sex + spl(std_density, k=5) + 
                           spl(std_julian, k=3) + std_age,
                         random = ~squirrel_id + Year.f + animal,
                         prior = GenoPrior_bi,
                         family = "threshold",
                         pedigree = Ainv,
                         data = rattle_probability_final, 
                         nitt = 250000, burnin = 50000, thin = 200, verbose=T)

plot(RattleModel2)
summary(RattleModel2)

RattleRepeatability <- RattleModel2$VCV[,1] / 
  (RattleModel2$VCV[,1] + RattleModel2$VCV[,2] + 
     RattleModel2$VCV[,3] + RattleModel2$VCV[,4])
hist(RattleRepeatability)
posterior.mode(RattleRepeatability)

RattleHerit <- RattleModel2$VCV[,3] / 
  (RattleModel2$VCV[,1] + RattleModel2$VCV[,2] + 
     RattleModel2$VCV[,3] + RattleModel2$VCV[,4])
hist(RattleHerit)
posterior.mode(RattleHerit)
```

### Intrusion Count

```{r, fig.height=8, fig.width=8}
IntrusionModel2 <- MCMCglmm(intruded ~ Grid + Sex + spl(std_density,k=5) + 
                           spl(std_julian,k=5) + std_age + std_local.density,
                            random = ~squirrel_id + Year.f + animal,
                            prior = GenoPrior_bi,
                            pedigree = Ainv,
                            family = "threshold",
                            data = intruders_model, 
                            nitt = 250000, burnin = 50000, thin = 200, verbose=F)

plot(IntrusionModel2)
summary(IntrusionModel2)

IntrusionRepeatability <- IntrusionModel2$VCV[,1] / 
  (IntrusionModel2$VCV[,1] + IntrusionModel2$VCV[,2] + 
     IntrusionModel2$VCV[,3] + IntrusionModel2$VCV[,4])
hist(IntrusionRepeatability)
posterior.mode(IntrusionRepeatability)

IntrusionHerit <- IntrusionModel2$VCV[,3] / 
  (IntrusionModel2$VCV[,1] + IntrusionModel2$VCV[,2] + 
     IntrusionModel2$VCV[,3] + IntrusionModel2$VCV[,4])
hist(IntrusionHerit)
posterior.mode(IntrusionHerit)
```

### Cache Size

```{r, fig.height=8, fig.width=8}
CacheSizeModel1 <- MCMCglmm(cache_size_lastyear ~ Grid + Sex + std_density + std_age +
                            I(std_age^2) + std_local_density,
                            random = ~squirrel_id + Year.f + animal,
                            prior = GenoPrior,
                            pedigree = Ainv,
                            family = "gaussian",
                            data = cache_sizes, 
                            nitt = 250000, burnin = 50000, thin = 200, verbose=F)


plot(CacheSizeModel2)
summary(CacheSizeModel2)

CacheSizeRepeatability <- CacheSizeModel2$VCV[,1] / 
  (CacheSizeModel2$VCV[,1] + CacheSizeModel2$VCV[,2] + 
     CacheSizeModel2$VCV[,3] + CacheSizeModel2$VCV[,4])
hist(CacheSizeRepeatability)
posterior.mode(CacheSizeRepeatability)

CacheSizeHerit <- CacheSizeModel2$VCV[,3] / 
  (CacheSizeModel2$VCV[,1] + CacheSizeModel2$VCV[,2] + 
     CacheSizeModel2$VCV[,3] + CacheSizeModel2$VCV[,4])
hist(CacheSizeHerit)
posterior.mode(CacheSizeHerit)
```

## Multivariate Phenotypic Models

Set the prior

```{r, eval=F}
MultiPhenoPrior <- list(G=list(G1=list(V = diag(4), nu = 4),
                               G2=list(V = diag(4), nu = 4)),
                        R=list(V = diag(4) , nu = 4))
```

We're using relatively informative priors with the same structure as a parameter expanded prior with a diagonal matrix of size 4 to match our number of traits.

```{r, eval=F}
MultiPhenoModel <- MCMCglmm(cbind(area, rattled, intrusions, cache_size)~
                              trait - 1 +
                              trait:Grid + trait:Sex + 
                              trait:spr_density + trait:log(total_events)+
                              trait:log(reflo_totals),
                            random = ~us(trait):squirrel_id +
                              us(trait):year.f,
                            
                            rcov = ~us(trait):units,
                            family = rep("poisson",4),
                            prior = MultiPhenoPrior,
                            data = Territoriality,
                            nitt = 2000000,
                            burnin = 1000000,
                            thin = 1000,
                            verbose=F)
```

## Multivariate Genetic Models

Set the prior

```{r, eval=F}
MultiGenoPrior <- list(G=list(G1=list(V = diag(4), nu = 4),
                              G2=list(V = diag(4), nu = 4),
                              G3=list(V = diag(4), nu = 4)),
                       R=list(V = diag(4) , nu = 4))
```

We're using relatively informative priors with the same structure as a parameter expanded prior with a diagonal matrix of size 4 to match our number of traits.

```{r, eval=F}
MultiGenoModel <- MCMCglmm(cbind(area, rattled, intrusions, cache_size)~
                             trait - 1 +
                             trait:Grid + 
                             trait:Sex + 
                             trait:spr_density + 
                             trait:log(total_events)+
                             trait:log(reflo_totals),
                           random = ~us(trait):squirrel_id + us(trait):animal + 
                             us(trait):year.f,
                           pedigree = Ainv,
                           rcov = ~us(trait):units,
                           family = rep("poisson",4),
                           prior = MultiGenoPrior,
                           data = Territoriality,
                           nitt = 2000000,
                           burnin = 1000000,
                           thin = 1000,
                           verbose=F)
```